### 2016-12-07 Projects/DelCo_2012/Report/DelCo_2012.tex
fix front matter caf
Minor changes to wording to remove inconsistencies (up to Section 2.3)

\begingroup
This publication is available free of charge from:
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1927}{http://dx.doi.org/10.6028/NIST.TN.XXXX} \\
\endgroup

@@

\begin{tabbing}
\hspace{1.5in} \= \\
BDP \> Bi-Directional Probe \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
DHS \> Department of Homeland Security \\
FEMA \> Federal Emergency Management Administration \\
HF \> Heat Flux \\
IDLH \> Immediate Danger to Life and Health \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
OSB \> Oriented Strand Board \\
TC \> Thermocouple \\
TFT \> Task Force Tips \\
UL FSRI \> UL Firefighter Safety Research Institute \\

@@

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both agents were effective in reducing the gas temperatures and suppressing fires in a compartment. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

This study, conducted by the National Institute of Standards and Technology (NIST), was part of a larger research project led by California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service.  This study was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

@@

A number of fire departments throughout the US have acquired CAFS units for structural fire fighting. Many have limited or stopped the use of these CAFS for structural fire fighting. This is due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural fire fighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural fire fighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide input on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural fire fighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A aqueous solution into an aerated form (a blanket of bubbles) that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or CAF can be used for an exterior (indirect) fire suppression attack on a structure.

@@

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. That study examined a broad range of tests to determine those parameters which most critically effect fire fighting performance. The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the aqueous solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate adsorbent fuels. These enhancements also make the agents valuable tools for mop-up or overhaul. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically an environment that is not an immediate danger to life and health (non-IDLH environment). In structural fire fighting, crews typically enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer and or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order and the experience of the MCFRS played a major role in the development of this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The experiments were planned during the workshop~\cite{Grant:2011}. Three types of experiments examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The logic was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. Wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets, carpeting, and upholstered furniture.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads. Project partner MCFRS provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1 \textregistered, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 22~mm (0.88~in) slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 30.5~m (100~ft) of 44~mm (1.75~in) diameter hose. Nozzles were mounted to a Blitz-Fire \textregistered monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

Thirty-two tests were conducted to determine spray density measures based on various nozzle patterns and locations. The gas cooling and fire suppression tests simulated an interior fire attack through the deployment of hose lines inside the structures with fixed nozzles. The purpose of these preliminary tests was to determine the appropriate nozzle locations for the gas cooling and fire suppression tests to follow by gaining an understanding of the water distribution throughout the structures.

Twenty-one of the tests were conducted in the concrete burn building where the gas cooling tests would take place, and twelve of the tests were conducted in the experimental structures to be used for the fire suppression tests. The test variables were the nozzle height, nozzle location within the test structure, and the nozzle pattern (Figure~\ref{fig:Spray_Density_Nozzle_Patterns}). A description and drawings of these facilities are included in Section~\ref{sec:Experimental Structures}. Detailed discussion of the tests and results are in Section~\ref{sec:Spray_Density}.

@@

Eighty-eight experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room. Suppression agents were not applied to the fire in the burn room.  Detailed discussion of the tests and results are in Section~\ref{sec:Gas_Cooling}.

@@

The fire suppression series included 20 experiments. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. Six experiments were conducted in the lower level of the two-story structure using both water and CAF from a straight stream nozzle.  In addition, two CAF experiments were conducted in the two-story burn building with a solid stream. The lower level of the two story structure allowed for a fire compartment that was approximately twice as large as the fire compartment in the single story.

@@

The series of tests described within this report were conducted at the Delaware County ESTC in Sharon Hill, PA. A burn building and two purpose-built concrete structures are located on the grounds of the ESTC. The burn building was used for the gas cooling experiments, and the purpose-built concrete structures were used for the fire suppression experiments. Figure~\ref{fig:Door_Status_Legend} provides the key used to identify doors and windows in the facility schematics.

@@

The burn building, or live fire training structure, contained a two-story section and a three-story section. The building was supported with reinforced concrete beams and columns. The floors and ceilings were also concrete, and the interior walls were constructed of cement block. The walls and ceiling of the burn rooms were protected with a 25~mm (1~in) thick layer of calcium silicate insulation and was covered by 50~mm (2~in) thick concrete tile. The floors of the burn rooms were protected with fire brick. Tests were conducted on the ground floor of the two-story, two-room configuration, which had overall dimensions of 6.5~m (21.3~ft) by 9.2~m (30.2~ft). Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout} contains a basic plan view of the test floor. A fully dimensioned plan view of the floor is presented in Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout_Detailed} in Appendix~\ref{app:floor_plans}.

@@

The burn room where the fuel load was located measured 3.8~m (12.5~ft) by 5.7~m (18.7~ft) with a ceiling height of 3.35~m (11.0~ft). The adjacent room where the water was applied for gas cooling measured 4.1~m (13.4~ft) by 5.7~m (18.7~ft), also with a ceiling height of 3.35~m (11.0~ft). The open doorway that connected the burn room to the adjacent room and the doorway from the adjacent room to the exterior both measured 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide.

@@

Two identical single-story concrete structures were built on a concrete slab as shown in Figure~\ref{fig:Delaware_County,_PA_Fire_Test_Structures}. They were designed to simulate a single floor of a residential structure.  The outer wall of each structure was composed of interlocking concrete blocks 0.61~m (2~ft) wide, 0.61~m (2~ft) high and 1.22~m (4~ft) long.  The interior dimensions of each structure were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. The joints and gaps between the blocks were filled with high temperature insulation.

@@

Each structure had a burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.4~m (8.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A basic plan view of the structure is presented in Figure~\ref{fig:Test_Structure_Floor_Plan}, and a fully dimensioned plan view of the structure is included in in Appendix~\ref{app:floor_plans}.

@@

The floor of the structure was the concrete pad. The interior walls of the structure were framed with steel studs and track. The studs were set to 0.40~m (16~in) centers. The ceiling support was composed of wood truss joist I-beams (TJIs) with a 299~mm (11.75~in) depth. The TJI was composed of laminated veneer lumber flanges with a cross section of 29~mm (1.13~in) x 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board web. Tongue and grove, 18.3~mm (0.72~in) thick, oriented strand board was attached to the top of the TJIs.

The interior walls of the burn room were lined with 13~mm (0.5~in) thick cement board. The ceiling of the burn room was the exposed ``floor assembly''. The walls of the hallway and entry foyer were composed of 16~mm (0.63~in) Type X gypsum board. The ceiling of the hallway and entry foyer was composed of two layers of 13~mm (0.5~in) thick cement board.

@@

A second story was added to the west single-story structure in order to conduct suppression experiments in a larger fire compartment than the fire compartment in the single story and to examine the impact of the fire and suppression in a remote compartment (upstairs) that was not directly exposed to the hose stream (see Figure~\ref{fig:delco_2story}).

@@

The outer walls of the bottom floor of the two-story structure were identical to that of the single-story structures described above but with an open floor plan. The interior dimensions were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. Figure~\ref{fig:dimensioned_first_2story} shows a basic floor plan of the reconfigured first floor. A fully dimensioned plan view of the floor is included in Appendix~\ref{app:floor_plans}. A stairwell was built to connect the two floors of the structure. The stairs had a 180~mm (7.25~in) rise and 190~mm (7.5~in) run and started 1.6~m (5.25~ft) off the south wall with a width of 1.2~m (4~ft) off the east wall. The second story walls were wood framed with 51~mm (2~in) by 102~mm (4~in) studs. The studs were set to 0.40~m (16~in) centers. The interior  walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) Durock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.31~in) oriented strand board and 8~mm (0.31~in) fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows a basic floor plan of the second story of the modified west structure, and a fully dimensioned plan view of the floor is included in Appendix~\ref{app:floor_plans}.

@@

The pallet stacks were located 0.3~m (12~in) away from the east and south walls of the burn room (see Figure~\ref{fig:Wood_Fuel_Load_Dimensions}). The pallets were 1.2~m (48~in) long, 1.0~m (40~in) wide, and 123~mm (4.8~in) high. The stacks were spaced 150~mm (6~in) apart. One layer of 13~mm (0.5~in) thick gypsum board panels were laid on the concrete floor under the wood pallets to form a protective layer to minimize thermal damage to the concrete floor. The east and south walls of the burn room were covered with 15~mm (0.59~in) thick sheets of plywood.  The wood ``flooring assembly'', composed of 12 TJIs and nine sheets of oriented strand board (OSB), served as the ceiling of the burn room. The open doorway to the burn room on the south side was covered with a 2.4~m (8~ft) tall by 1.2~m (4~ft) wide piece of medium density fiberboard paneling that was 5~mm (0.19~in) thick.

>>>>

\begingroup
This publication is available free of charge from:
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1927}{http://dx.doi.org/10.6028/NIST.TN.1927} \\
\endgroup

@@

\begin{tabbing}
\hspace{1.5in} \= \\
BDP \> Bi-Directional Probe \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
DHS \> Department of Homeland Security \\
FEMA \> Federal Emergency Management Administration \\
HF \> Heat Flux \\
IDLH \> Immediate Danger to Life and Health \\
ESTC \> Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
OSB \> Oriented Strand Board \\
TC \> Thermocouple \\
% TFT \> Task Force Tips \\
% UL FSRI \> Underwriters Laboratories Firefighter Safety Research Institute \\

@@

Full scale fire experiments designed to address the key outcomes were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both agents were effective in reducing the gas temperatures and suppressing fires in a compartment. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

This study, conducted by the National Institute of Standards and Technology, was part of a larger research project led by California Polytechnic State University, San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service. This study was supported in part by a Department of Homeland Security/Federal Emergency Management Administration Assistance to Firefighters Research and Development Grant.

@@

A number of fire departments throughout the United States have acquired CAFS units for structural fire fighting. Many have limited or stopped the use of these CAFS for structural fire fighting. This is due to operational challenges or due to lack of understanding related to fire ground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural fire fighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural fire fighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide input on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural fire fighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution; nozzle reaction force; the force to kink the hose; potential for slipping on a variety of surfaces; and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A aqueous solution into an aerated form (a blanket of bubbles) that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they have already applied the agent, allowing them to make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire, or CAF can be used for an exterior (indirect) fire suppression attack on a structure.

@@

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. The study examined a broad range of tests to determine the parameters that most critically affect fire fighting performance. The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the aqueous solutions provided enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls, and the reduced inter-facial surface tension allowed the water to penetrate adsorbent fuels. These enhancements make the agents valuable tools for mop-up or overhaul. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

In 2005, the National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified during the event was ``improved extinguishing agents and delivery systems''. The report from the symposium states that ``the primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems''. In wildland fire fighting, CAF is typically applied directly to the fuel from a remote location, typically an environment that is not an immediate danger to life and health (non-IDLH environment). In structural fire fighting, crews typically enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer and/or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order and the experience of the MCFRS played a major role in the development of this study.

@@

All experiments described in this report were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The experiments were planned during a workshop with the interested parties~\cite{Grant:2011}. Three types of experiments examining spray density measures, gas layer cooling, and fire suppression of fuel packages containing upholstered furniture were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The idea was that the synthetic materials in the upholstered furniture burned only on the surface and thus were easily extinguished. Wood-based fuels, however, would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that the following additional experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained wood-based wall and ceiling finishes; wooden pallets; carpeting; and upholstered furniture.

The following sections describe the various types of experiments, facilities, instrumentation, and fuel loads used during the tests. Project partner MCFRS provided the CAFS equipped engines along with the necessary staffing to provide both water and CAF for the experiments. All experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression. Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips Metro 1~\textregistered, fixed gallonage nozzles with a flow rate of 120~gpm. These were automatic nozzles with a 22~mm (0.88~in) slug-tip and were able to be adjusted between a straight stream and a wide fog stream for application. The setup also included 30.5~m (100~ft) of 44~mm (1.75~in) diameter hose. Nozzles were mounted to a Blitz-Fire~\textregistered~ monitor device and placed inside the structures. The nozzles were controlled and operated from the exterior via a gate valve. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye was adjusted to allow the CAF to flow into the test structure. This technique minimized the build-up of ``dry'' foam on the slug upon the initial flow of the nozzle into the test structure.

@@

Thirty-two tests were conducted to determine spray density measures based on various nozzle patterns and locations. The purpose of these preliminary tests was to determine the appropriate nozzle locations for the gas cooling and fire suppression tests by gaining an understanding of the water distribution throughout the structures. The gas cooling and fire suppression tests simulated an interior fire attack through the deployment of hose lines inside the structures with fixed nozzles.

Twenty-one of the spray density tests were conducted in the concrete burn building that was later used for the gas cooling tests, and twelve of the tests were conducted in the experimental structures that were later used for the fire suppression tests. The test variables were the nozzle height, nozzle location within the test structure, and the nozzle pattern (Figure~\ref{fig:Spray_Density_Nozzle_Patterns}). Descriptions of these facilities and their floor plan schematics are included in Section~\ref{sec:Experimental Structures}. A detailed discussion of the spray density tests and results is included in Section~\ref{sec:Spray_Density}.

@@

Eighty-eight experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room. Suppression agents were not applied to the fire in the burn room. A detailed discussion of the gas cooling tests and results is included in Section~\ref{sec:Gas_Cooling}.

@@

The fire suppression series included 20 experiments. Twelve experiments were conducted in the single-story structures (six in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. Six experiments were conducted in the lower level of the two-story structure using both water and CAF from a straight stream nozzle. Additionally, two CAF experiments were conducted in the two-story burn building with a solid stream. The lower level of the two-story structure allowed for a fire compartment that was approximately twice as large as the fire compartment in the single-story.

@@

The series of tests described within this report were conducted at the Delaware County ESTC in Sharon Hill, PA. A burn building and two purpose-built concrete structures were located on the grounds of the ESTC. The burn building was used for the gas cooling experiments, and the purpose-built concrete structures were used for the fire suppression experiments. Various floor plan schematics of the structures are included in the following sections. Figure~\ref{fig:Door_Status_Legend} provides the key used to identify doors and windows in the schematics.

@@

The burn building, or live fire training structure, contained a two-story section and a three-story section. The building was supported with reinforced concrete beams and columns. The floors and ceilings were also concrete, and the interior walls were constructed of cement block. The walls and ceiling of the burn rooms were protected with a 25~mm (1~in) thick layer of calcium silicate insulation and was covered by 50~mm (2~in) thick concrete tile. The floors of the burn rooms were protected with fire brick. Tests were conducted in a two-room configuration located on the ground floor of the two-story section. The overall dimensions of the test floor were 6.5~m (21.3~ft) by 9.2~m (30.2~ft). Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout} contains a basic plan view of the two-room configuration, and a fully dimensioned plan view of the test floor is presented in Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout_Detailed} in Appendix~\ref{app:floor_plans}.

@@

The fuel load was located in the burn room, which had interior dimensions of 3.8~m (12.5~ft) by 5.7~m (18.7~ft) and a ceiling height of 3.4~m (11.0~ft). Water application for the purpose of gas cooling always occurred in the adjacent room, which had interior dimensions of 4.1~m (13.4~ft) by 5.7~m (18.7~ft) and a ceiling height of 3.4~m (11.0~ft). The open doorway that connected the burn room to the gas cooling room and the doorway from the gas cooling room to the exterior both measured 2.0~m (6.5~ft) high and 0.9~m (3.0~ft) wide.

@@

Two identical single-story concrete structures were built on a concrete slab as shown in Figure~\ref{fig:Delaware_County,_PA_Fire_Test_Structures}. They were designed to simulate a single floor of a residential structure. The outer wall of each structure was composed of interlocking concrete blocks 0.6~m (2.0~ft) wide, 0.6~m (2.0~ft) high and 1.2~m (4.0~ft) long. The interior dimensions of each structure were 6.1~m (20.0~ft) wide, 11.0~m (36.0~ft) long, and 2.4~m (8.0~ft) high. The joints and gaps between the blocks were filled with high temperature insulation.

@@

Each structure had a burn room with interior dimensions of approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) by 2.4~m (8.0~ft) high. A 5.4~m (17.6~ft) long hallway with a width of 0.9~m (3.1~ft) and ceiling height of 2.4~m (8.0~ft) connected the burn room to a small entry foyer at the front of the structure. The entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft) and had a ceiling height of 2.4~m (8.0~ft). A 2.0~m (6.5~ft) high by 0.9~m (3.0~ft) wide open doorway on the north side of the structure connected the entry foyer to the exterior. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A basic plan view of the structure is presented in Figure~\ref{fig:Test_Structure_Floor_Plan}, and a fully dimensioned plan view is included in Figure~\ref{fig:Test_Structure_Floor_Plan_Detailed} in Appendix~\ref{app:floor_plans}.

@@

The floor of the structure was the concrete pad. The interior walls of the structure were framed with steel studs and track. The studs were set to 0.4~m (16~in) centers. The ceiling support was composed of wood truss joist I-beams (TJIs) with a 0.3~m (11.75~in) depth. Each TJI was composed of laminated veneer lumber flanges with a cross section of 29~mm (1.13~in) x 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board (OSB) web. Tongue and grove, 18~mm (0.72~in) thick, OSB was attached to the top of the TJIs.

The interior walls of the burn room were lined with 13~mm (0.5~in) thick cement board. The ceiling of the burn room was the exposed wood TJIs from the structure's ceiling support. The walls of the hallway and entry foyer were composed of 16~mm (0.63~in) Type X gypsum board. The ceiling of the hallway and entry foyer was composed of two layers of 13~mm (0.5~in) thick cement board.

@@

A second story was added to the west single-story structure in order to conduct suppression experiments in a fire compartment larger than the single-story fire compartment and to examine the impact of the fire and suppression on a remote compartment (the second floor) not directly exposed to the hose stream (see Figure~\ref{fig:delco_2story}).

@@

The outer walls of the bottom floor of the two-story structure were identical to the outer walls of the single-story structures described above, but the interior layout of the bottom floor was an open floor plan. The large, open space on the interior of the ground floor was approximately 6.1~m (20.0~ft) wide, 11.0~m (36.0~ft) long, and 2.4~m (8.0~ft) high. A basic plan view of the reconfigured floor is presented in Figure~\ref{fig:dimensioned_first_2story}, and a fully dimensioned plan view is included in Figure~\ref{fig:dimensioned_first_2story_detailed} in Appendix~\ref{app:floor_plans}.

A set of stairs, each with a 180~mm (7.25~in) rise and 190~mm (7.5~in) run, was built along the interior of the east wall to connect the two floors of the structure. The stairway was 1.2~m (4.0~ft) wide, and the edge of the bottom step was located 1.6~m (5.25~ft) from the south wall.The second story walls were wood framed with 51~mm (2.0~in) by 102~mm (4.0~in) studs. The studs were set to 0.40~m (16~in) centers. The interior walls were protected by a layer of 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) Durock cement board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected by 11~mm (0.31~in) thick OSB and 8~mm (0.31~in) thick fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows a basic plan view of the second story of the modified west structure, and a fully dimensioned plan view of the floor is provided in Figure~\ref{fig:dimensioned_second_2story_detailed} in Appendix~\ref{app:floor_plans}.

@@

The pallet stacks were located 0.3~m (12~in) away from the east and south walls of the burn room (see Figure~\ref{fig:Wood_Fuel_Load_Dimensions}). The pallets were 1.2~m (48~in) long, 1.0~m (40~in) wide, and 123~mm (4.8~in) high. The stacks were spaced 150~mm (6~in) apart. One layer of 13~mm (0.5~in) thick gypsum board panels were laid on the concrete floor under the wood pallets to form a protective layer to minimize thermal damage to the concrete floor. The east and south walls of the burn room were covered with 15~mm (0.59~in) thick sheets of plywood.  The wood ``flooring assembly'', composed of 12 TJIs and nine sheets of OSB, served as the ceiling of the burn room. The open doorway to the burn room on the south side was covered with a 2.4~m (8~ft) tall by 1.2~m (4~ft) wide piece of medium density fiberboard paneling that was 5~mm (0.19~in) thick.
